---
title: "Classification Warning Example"
output: github_document
---


# [`vtreat`](https://github.com/WinVector/vtreat) Nested Model Bias Warning

For quite a while we have been teaching estimating variable re-encodings on the exact same data they
are later *naively* using to train a model on leads to an undesirable nested model bias.  The `vtreat`
package (both the [`R` version](https://github.com/WinVector/vtreat) and 
[`Python` version](https://github.com/WinVector/pyvtreat)) both incorporate a cross-frame method
that allows one to use all the training data both to build learn variable re-encodings and to correctly train a subsequent model (for an example please see our recent [PyData LA talk](http://www.win-vector.com/blog/2019/12/pydata-los-angeles-2019-talk-preparing-messy-real-world-data-for-supervised-machine-learning/)).

The next version of `vtreat` will warn the user if they have improperly used the same data for both `vtreat` impact code inference and downstream modeling.  So in addition to us warning you not to do this, the package now also checks and warns against this situation.

## Set up the Example

This example is copied from [some of our classification documentation](https://github.com/WinVector/vtreat/blob/master/Examples/Classification/Classification.md).


Load modules/packages.

```{r}
# For this example we want vtreat version 1.5.1 or newer
# remotes::install_github("WinVector/vtreat")
library(vtreat)

packageVersion("vtreat")
```

Generate example data. 

* `y` is a noisy sinusoidal function of the variable `x`
* `yc` is the output to be predicted: whether `y` is > 0.5. 
* Input `xc` is a categorical variable that represents a discretization of `y`, along some `NA`s
* Input `x2` is a pure noise variable with no relationship to the output

```{r}
make_data <- function(nrows) {
    d <- data.frame(x = 5*rnorm(nrows))
    d['y'] = sin(d['x']) + 0.1*rnorm(n = nrows)
    d[4:10, 'x'] = NA                  # introduce NAs
    d['xc'] = paste0('level_', 5*round(d$y/5, 1))
    d['x2'] = rnorm(n = nrows)
    d[d['xc']=='level_-1', 'xc'] = NA  # introduce a NA level
    d['yc'] = d[['y']]>0.5
    return(d)
}

training_data = make_data(500)

training_data %.>%
  head(.) %.>%
  knitr::kable(.)
```


## Demonstrate the Warning

Now that we have the data, we want to treat it prior to modeling: we want training data where all the input variables are numeric and have no missing values or `NA`s.

First create the data treatment transform design object, in this case a treatment for a binomial classification problem.

We use the training data `training_data` to fit the transform and the return a treated training set: completely numeric, with no missing values.

```{r}
transform_design = vtreat::mkCrossFrameCExperiment(
    dframe = training_data,                                    # data to learn transform from
    varlist = setdiff(colnames(training_data), c('y', 'yc')),  # columns to transform
    outcomename = 'yc',                            # outcome variable
    outcometarget = TRUE                           # outcome of interest
)
transform <- transform_design$treatments
train_prepared <- transform_design$crossFrame
```

`train_prepared` is prepared in the correct way to use the same training data for inferring the impact-coded variables, using the returned `$crossFrame` from `mkCrossFrameCExperiment()`.

We prepare new test or application data as follows.

```{r}
test_data <- make_data(100)

test_prepared <- prepare(transform, test_data)
```

The issue is: for training data we should not call `prepare()`, but instead use the cross-frame
that is produces during transform design.

The point is we should not do the following:

```{r}
train_prepared_wrong <- prepare(transform, training_data)
```

Notice we now get a warning that we should not have done this, and in doing so we may have a nested model bias data leak.

And that is the new nested model bias warning feature.

The `Python`-version of this document can be found [here](https://github.com/WinVector/pyvtreat/blob/master/Examples/Classification/ClassificationWarningExample.md).

