---
title: "Multi Class vtreat"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi Class vtreat}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Problem: try to prepare data to model multi-class y as a function of x using [`vtreat`](https://github.com/WinVector/vtreat). `vtreat` does not directly do this, but can be used to do this.  In this note we share functions to make this adaption.

```{r libs}
library("vtreat")
```

The two functions needed ([`mkCrossFrameMExperiment()`](https://winvector.github.io/vtreat/reference/mkCrossFrameMExperiment.html) and [`prepare_m()`](https://winvector.github.io/vtreat/reference/prepare_m.html)) are (as of version `1.2.4`) part of `vtreat`.
For earlier versions of `vtreat` you can copy them into your own work.

```{r fns, comment=""}
print(mkCrossFrameMExperiment)

print(prepare_m)
```

Our specific example: try to model multi-class y as a function of x1 and x2

```{r mkex}
# create example data
set.seed(326346)
sym_bonuses <- rnorm(3)
names(sym_bonuses) <- c("a", "b", "c")
n_row <- 1000
d <- data.frame(x1 = rnorm(n_row),
                x2 = sample(names(sym_bonuses), n_row, replace = TRUE),
                y = "NoInfo",
                stringsAsFactors = FALSE)
d$y[d$x1 > sym_bonuses[d$x2] + runif(n_row)] <- "Large1"
d$y[sym_bonuses[d$x2] > d$x1 + runif(n_row)] <- "Large2"

knitr::kable(head(d))
```

We define the problem controls and use `mkCrossFrameMExperiment()` to build both a cross-frame and
a treatment plan.

```{r tdef}
# define problem
vars <- c("x1", "x2")
y_name <- "y"
y_levels <- sort(unique(d[[y_name]]))

# build the multi-class cross frame and treatments
cfe_m <- mkCrossFrameMExperiment(d, vars, y_name)
```

The cross-frame is the entity safest for training on (unless you have made separate data split for the 
treatment design step). It uses cross-validation to reduce nested model bias. Some notes on this issue
are available [here](https://winvector.github.io/vtreat/articles/vtreatCrossFrames.html), and [here](https://github.com/WinVector/vtreat/blob/master/extras/vtreat.pdf).

```{r crossframe}
# look at the data we would train models on
str(cfe_m$cross_frame)
```

`prepare_m()` is a function that can apply the designed treatments to new data.  Here we are simulating new 
data by re-using our design data.

```{r treatment plan}
# pretend original data is new data to be treated
# NA out top row to show processing
for(vi in vars) {
  d[[vi]][[1]] <- NA
}
str(prepare_m(d, cfe_m$treat_m))
```

Obvious issues include: computing variable significances, and blow up and co-dependency of produced columns.  These we leave for the next modeling step to deal with (this is our philosophy with most issues that involve joint distributions of variables).


