---
title: "Regression Fit Prepare Notation"
output: github_document
---


# Using vtreat with Regression Problems 


Nina Zumel and John Mount
February 2020

This article documents `vtreat`'s ["fit_prepare" variation](https://github.com/WinVector/vtreat/blob/master/Examples/fit_transform/fit_prepare_api.md) for regression problems.
This API was inspired by the [`pyvtreat`](https://github.com/WinVector/pyvtreat) API, which was in turn based on the `.fit()`, `.transform()`, `.fit_transform()` workflow of `scikit-learn` in `Python`. 

The same example in the original `R` `vtreat` notation can be found [here](https://github.com/WinVector/vtreat/blob/master/Examples/Regression/Regression.md). 

The same example in the [`Python` version of `vtreat`](https://github.com/WinVector/pyvtreat) can be found [here](https://github.com/WinVector/pyvtreat/blob/master/Examples/Regression/Regression.md).


## Preliminaries

Load modules/packages.

```{r}
library(vtreat)
packageVersion('vtreat')
suppressPackageStartupMessages(library(ggplot2))
library(WVPlots)
library(rqdatatable)
```

Generate example data. 

* `y` is a noisy sinusoidal plus linear function of the variable `x`, and is the output to be predicted
* Input `xc` is a categorical variable that represents a discretization of `y`, along with some `NaN`s
* Input `x2` is a pure noise variable with no relationship to the output

```{r}
set.seed(2020)

make_data <- function(nrows) {
    d <- data.frame(x = 5*rnorm(nrows))
    d['y'] = sin(d[['x']]) + 0.01*d[['x']] + 0.1*rnorm(n = nrows)
    d[4:10, 'x'] = NA                  # introduce NAs
    d['xc'] = paste0('level_', 5*round(d$y/5, 1))
    d['x2'] = rnorm(n = nrows)
    d[d['xc']=='level_-1', 'xc'] = NA  # introduce a NA level
    return(d)
}

d = make_data(500)

d %.>%
  head(.) %.>%
  knitr::kable(.)
```

### Some quick data exploration


Check how many levels `xc` has, and their distribution (including `NaN`)


```{r}
unique(d['xc'])
```

```{r}
table(d$xc, useNA = 'always')
```


Find the mean value of `y`


```{r}
mean(d[['y']])
```


Plot of `y` versus `x`.


```{r}
ggplot(d, aes(x=x, y=as.numeric(y))) + 
  geom_line()
```

## Build a transform appropriate for regression problems.

Now that we have the data, we want to treat it prior to modeling: we want training data where all the input variables are numeric and have no missing values or `NaN`s.

First create the data treatment transform object, in this case a treatment for a regression problem.

```{r}
transform_design <- vtreat::NumericOutcomeTreatment(
    var_list = setdiff(colnames(d), c('y')),  # columns to transform
    outcome_name = 'y'                        # outcome variable
)

```

Now call the `fit_prepare()` function with the training data `d` to fit the transform and also return a treated training set. The `fit_prepare()` function returns the fitted data treatment object (as `treatments`) and a
statistically correct treated training set (as `cross_frame`) for training the model. The `cross_frame` is guaranteed to be completely numeric, with no missing values.


```{r}
# the unpack notation is a multiassignment operator
# see https://winvector.github.io/wrapr/articles/unpack_multiple_assignment.html
# for more details

unpack[treatment_plan = treatments,
       d_prepared = cross_frame] <- fit_prepare(transform_design, d)

# list the derived variables
get_feature_names(treatment_plan)

```

Notice that `d_prepared` only includes derived variables and the outcome `y`. We will discuss the types of derived variables below.

```{r}
d_prepared %.>%
  head(.) %.>%
  knitr::kable(.)
```

As we will see below, the `prepare()` function applies the fitted data treatments to future data, prior to calling your model on the data. Note that for the training data `d`: `fit_prepare(transform_spec, d)` is **not** the same as
`fit(transform_spec, d) %.>% prepare(., d)`; the second call can lead to nested model bias in some
situations, and is **not** recommended. In other words, it is a bad idea to call `prepare()` on your original training data.

For future application data `df` that is not seen during transform design, `prepare(treatment_plan, df)` is the appropriate step.

`vtreat` version `1.5.1` and newer issue a warning if you call the incorrect transform pattern on your original training data:

```{r}
d_prepared_wrong <- prepare(treatment_plan, d)
```


## The Score Frame

Now examine the score frame, which gives information about each new variable, including its type, which original variable it is derived from, its (cross-validated) significance as a one-variable linear model for the outcome,and the (cross-validated) R-squared of its corresponding linear model.

```{r}
# get statistics on the variables
score_frame <- get_score_frame(treatment_plan)

# only print a subset of the columns
cols = c("varName", "origName", "code", "rsq", "sig", "varMoves", "default_threshold", "recommended")
knitr::kable(score_frame[,cols])
```

Notice that the variable `xc` has been converted to multiple variables: 

* an indicator variable for each common possible level (`xc_lev_*`)
* the value of a (cross-validated) one-variable model for `y` as a function of `xc` (`xc_catN`)
* a variable that returns how prevalent this particular value of `xc` is in the training data (`xc_catP`)
* a variable that returns the standard deviation of `y` conditioned on `xc` (`xc_catD`)

The variable `x` has been converted to two variables:

* a clean version of `x` that has no missing values of `NaN`s
* a variable indicating when `x` was `NaN` or `NA` in the original data (`x_isBAD`)

Any or all of these new variables are available for downstream modeling.

The `recommended` column indicates which variables are non constant (`has_range` == True) and have a significance value smaller than `default_threshold`. See the section *Deriving the Default Thresholds* below for the reasoning behind the default thresholds. Recommended columns are intended as advice about which variables appear to be most likely to be useful in a downstream model. This advice attempts to be conservative, to reduce the possibility of mistakenly eliminating variables that may in fact be useful (although, obviously, it can still mistakenly eliminate variables that have a real but non-linear relationship to the output).

Let's look at the recommended and not recommended variables:

```{r}
# recommended variables
score_frame[score_frame[['recommended']], 'varName', drop = FALSE]  %.>%
  knitr::kable(.)
```

```{r}
# not recommended variables
score_frame[!score_frame[['recommended']], 'varName', drop = FALSE] %.>%
  knitr::kable(.)
```

## A Closer Look at the `catN` variables

Variables of type `catN` are the outputs of a one-variable hierarchical linear regression of a categorical variable (in our example, `xc`) against the centered output on the (cross-validated) treated training data. 

Let's look at the relationship between `xc_catN` and `y` (actually `y_centered`, a centered version of `y`). 

```{r}
d_prepared['y_centered'] <- d_prepared$y - mean(d_prepared$y)

WVPlots::ScatterHist(
  d_prepared, 
  xvar = 'xc_catN',
  yvar = 'y_centered',
  smoothmethod = 'identity',
  estimate_sig = TRUE,
  title = 'Relationship between xc_catN and y')
```

This indicates that `xc_catN` is strongly predictive of the outcome. Note that the score frame also reported the R-squared between  `xc_catN` and `y_centered`, which is fairly large.

```{r}
score_frame[score_frame$varName == 'xc_catN', ]$rsq
```

Note also that the impact code values are jittered; this is because `d_prepared` is a "cross-frame": that is, the result of a cross-validated estimation process. Hence, the impact coding of `xc` is a function of both the value of `xc` and the cross-validation fold of the datum's row. When `transform` is applied to new data, there will be only one value of impact code for each (common) level of `xc`. We can see this by applying the transform to the data frame `d` as if it were new data.


```{r}
# the scores for the rows in the cross_frame where xc == 'level_1'
# jittered
summary(d_prepared$xc_catN[(!is.na(d$xc)) & (d$xc == 'level_1')])

# notice this generates a warning!
dtmp = prepare(treatment_plan, d)

dtmp['y_centered'] = dtmp$y - mean(dtmp$y)

# the scores for the rows of "new" prepared data where xc == 'level_1'
# constant
summary(dtmp$xc_catN[(!is.na(d$xc)) & (d$xc == 'level_1')])
```

Variables of type `catN` are useful when dealing with categorical variables with a very large number of possible levels. For example, a categorical variable with 10,000 possible values potentially converts to 10,000 indicator variables, which may be unwieldy for some modeling methods. Using a single numerical variable of type `catN` may be a preferable alternative.

## Using the Prepared Data in a Model

Of course, what we really want to do with the prepared training data is to fit a model jointly with all the (recommended) variables. 
Let's try fitting a linear regression model to `d_prepared`.


```{r}
model_vars <- score_frame$varName[score_frame$recommended]
# to use all the variables:
# model_vars <- score_frame$varName

f <- wrapr::mk_formula('y', model_vars)

model = lm(f, data = d_prepared)

# now predict
d_prepared['prediction'] = predict(
  model,
  newdata = d_prepared)

# look at the fit (on the training data)
WVPlots::ScatterHist(
  d_prepared, 
  xvar = 'prediction',
  yvar = 'y',
  smoothmethod = 'identity',
  estimate_sig = TRUE,
  title = 'Relationship between prediction and y')
```

Now apply the model to new data.

```{r}
# create the new data
dtest <- make_data(450)

# prepare the new data with vtreat
dtest_prepared = prepare(treatment_plan, dtest)

# apply the model to the prepared data
dtest_prepared['prediction'] = predict(
  model,
  newdata = dtest_prepared)

# compare the predictions to the outcome (on the test data)
WVPlots::ScatterHist(
  dtest_prepared, 
  xvar = 'prediction',
  yvar = 'y',
  smoothmethod = 'identity',
  estimate_sig = TRUE,
  title = 'Relationship between prediction and y')

# get r-squared
sigr::wrapFTest(dtest_prepared, 
                predictionColumnName = 'prediction',
                yColumnName = 'y',
                nParameters = length(model_vars) + 1)
```

## Parameters for `NumericOutcomeTreatment`

We've tried to set the defaults for all parameters so that `vtreat` is usable out of the box for most applications.


```{r}
regression_parameters()
```

Some parameters of note include:

**codeRestriction**: The types of synthetic variables that `vtreat` will (potentially) produce. By default, all possible applicable types will be produced. See *Types of prepared variables* below.

**minFraction** (default: 0.02): For categorical variables, indicator variables (type `indicator_code`) are only produced for levels that are present at least `minFraction` of the time. A consequence of this is that 1/`minFraction` is the maximum number of indicators that will be produced for a given categorical variable. To make sure that *all* possible indicator variables are produced, set `minFraction = 0`

**splitFunction**: The cross validation method used by `vtreat`. Most people won't have to change this.

**ncross** (default: 3): The number of folds to use for cross-validation

**missingness_imputation**: The function or value that vtreat uses to impute or "fill in" missing numerical values. The default is `mean`. To change the imputation function or use different functions/values for different columns, see the [Imputation example](https://github.com/WinVector/vtreat/blob/master/Examples/Imputation/Imputation_FP.md)


**customCoders**: For passing in user-defined transforms for custom data preparation. Won't be needed in most situations, but see [here](http://www.win-vector.com/blog/2017/09/custom-level-coding-in-vtreat/) for an example of applying a GAM transform to input variables.


## Types of prepared variables

**clean**: Produced from numerical variables: a clean numerical variable with no `NAs` or missing values

**lev**: Produced from categorical variables, one for each (common) level: for each level of the variable, indicates if that level was "on"

**catP**: Produced from categorical variables: indicates how often each level of the variable was "on" (its prevalence)

**catN**: Produced from categorical variables: score from a one-dimensional model of the centered output as a function of the explanatory variable

**catD**: Produced from categorical variables: deviation of outcome as a function of the explanatory variable

**isBAD**: Produced for numerical variables: an indicator variable that marks when the original variable was missing or `NaN`.

More on the coding types can be found [here](https://winvector.github.io/vtreat/articles/vtreatVariableTypes.html).

### Example: Produce only a subset of variable types

In this example, suppose you only want to use indicators and continuous variables in your model; 
in other words, you only want to use variables of types (`clean`, `is_BAD`, and `lev`), and no `catN`, `catP`, or `catD` variables.


```{r}
# create a new set of parameters, overriding 
# the default for codeRestriction
newparams = regression_parameters(
  list(
    codeRestriction = c('clean', 'isBAD', 'lev')
  ))

thin_spec <- vtreat::NumericOutcomeTreatment(
    var_list = setdiff(colnames(d), c('y')),  # columns to transform
    outcome_name = 'y',                       # outcome variable
    params = newparams                        # set the parameters 
)

unpack[
  thin_plan = treatments,
  thin_frame = cross_frame
  ] <- fit_prepare(thin_spec, d)   

# examine the new prepared training data
# no catX variables
knitr::kable(head(thin_frame))

# examine the score frame for the new treatment plan
# no catX variables
knitr::kable(get_score_frame(thin_plan)[,cols])

```


## Deriving the Default Thresholds

While machine learning algorithms are generally tolerant to a reasonable number of irrelevant or noise variables, too many irrelevant variables can lead to serious overfit; see [this article](http://www.win-vector.com/blog/2014/02/bad-bayes-an-example-of-why-you-need-hold-out-testing/) for an extreme example, one we call "Bad Bayes". The default threshold is an attempt to eliminate obviously irrelevant variables early.

Imagine that you have a pure noise dataset, where none of the *n* inputs are related to the output. If you treat each variable as a one-variable model for the output, and look at the significances of each model, these significance-values will be uniformly distributed in the range [0:1]. You want to pick a weakest possible significance threshold that eliminates as many noise variables as possible. A moment's thought should convince you that a threshold of *1/n* allows only one variable through, in expectation. 

This leads to the general-case heuristic that a significance threshold of *1/n* on your variables should allow only one irrelevant variable through, in expectation (along with all the relevant variables). Hence, *1/n* used to be our recommended threshold, when we originally developed the R version of `vtreat`.

We noticed, however, that this biases the filtering against numerical variables, since there are at most two derived variables (of types *clean* and *is_BAD*) for every numerical variable in the original data. Categorical variables, on the other hand, are expanded to many derived variables: several indicators (one for every common level), plus *catN*, *catP*, and *catD*. So we now reweight the thresholds.

Suppose you have a (treated) data set with *ntreat* different types of `vtreat` variables (`clean`, `lev`, etc). There are *nT* variables of type *T*. Then the default threshold for all the variables of type *T* is *1/(ntreat nT)*. This reweighting helps to reduce the bias against any particular type of variable. The heuristic is still that the set of recommended variables will allow at most one noise variable into the set of candidate variables.

As noted above, because `vtreat` estimates variable significances using linear methods by default, some variables with a non-linear relationship to the output may fail to pass the threshold. In this case, you may not wish to filter the variables to be used in the models to only recommended variables (as we did in the main example above), but instead use all the variables, or select the variables to use by your own criteria.


## Conclusion


In all cases (classification, regression, unsupervised, and multinomial classification) the intent is that `vtreat` transforms are essentially one liners.

The preparation commands are organized as follows:

 * **Regression**: [`R` regression example](https://github.com/WinVector/vtreat/blob/master/Examples/Regression/Regression_FP.md), [`Python` regression example](https://github.com/WinVector/pyvtreat/blob/master/Examples/Regression/Regression.md).
 * **Classification**: [`R` classification example](https://github.com/WinVector/vtreat/blob/master/Examples/Classification/Classification_FP.md), [`Python` classification  example](https://github.com/WinVector/pyvtreat/blob/master/Examples/Classification/Classification.md).
 * **Unsupervised tasks**: [`R` unsupervised example](https://github.com/WinVector/vtreat/blob/master/Examples/Unsupervised/Unsupervised_FP.md), [`Python` unsupervised example](https://github.com/WinVector/pyvtreat/blob/master/Examples/Unsupervised/Unsupervised.md).
 * **Multinomial classification**: [`R` multinomial classification example](https://github.com/WinVector/vtreat/blob/master/Examples/Multinomial/MultinomialExample_FP.md), [`Python` multinomial classification example](https://github.com/WinVector/pyvtreat/blob/master/Examples/Multinomial/MultinomialExample.md).

These current revisions of the examples are designed to be small, yet complete.  So as a set they have some overlap, but the user can rely mostly on a single example for a single task type.

