---
title: "Multinomial Classification Fit Transform Notation"
output: github_document
---

# Using [vtreat](https://github.com/WinVector/vtreat) with Multinomial Classification Problems

Nina Zumel and John Mount
October 2019

Note this the "fit_transform" variation (a notation closer to that used in [`pyvtreat`](https://github.com/WinVector/pyvtreat)) of the description of the [`R` version of `vtreat`](https://github.com/WinVector/vtreat). The original `vtreat` notation version can be found [here](https://github.com/WinVector/vtreat/blob/master/Examples/Multinomial/MultinomialExample.md). The same example for the [`Python` version of `vtreat`](https://github.com/WinVector/pyvtreat) can be found [here](https://github.com/WinVector/pyvtreat/blob/master/Examples/Multinomial/MultinomialExample.md).

## Preliminaries


Load modules/packages.

```{r}
library(rqdatatable)
library(vtreat)
suppressPackageStartupMessages(library(ggplot2))
library(WVPlots)
```

Generate example data. 

* `y` is a noisy sinusoidal function of the variable `x`
* `yc` is the output to be predicted: `y`'s quantized value as 'large', 'liminal', or 'small'.
* Input `xc` is a categorical variable that represents a discretization of `y`, along some `NA`s
* Input `x2` is a pure noise variable with no relationship to the output

```{r}
make_data <- function(nrows) {
    d <- data.frame(x = 5*rnorm(nrows))
    d['y'] = sin(d['x']) + 0.1*rnorm(n = nrows)
    d[4:10, 'x'] = NA                  # introduce NAs
    d['xc'] = paste0('level_', 5*round(d$y/5, 1))
    d['x2'] = rnorm(n = nrows)
    d[d['xc']=='level_-1', 'xc'] = NA  # introduce a NA level
    d['yc'] = ifelse(d[['y']] > 0.5, 'large', ifelse(d[['y']] < -0.5, 'small', 'liminal'))
    return(d)
}

d = make_data(500)

d %.>%
  head(.) %.>%
  knitr::kable(.)
```


### Some quick data exploration


Check how many levels `xc` has, and their distribution (including `NA`)

```{r}
unique(d['xc'])
```

```{r}
table(d$xc, useNA = 'ifany')
```

Show the distribution of `yc`

```{r}
table(d$yc, useNA = 'ifany')
```

## Build a transform appropriate for classification problems.

Now that we have the data, we want to treat it prior to modeling: we want training data where all the input variables are numeric and have no missing values or `NA`s.

First create the data treatment transform design object, in this case a treatment for a binomial classification problem.

We use the training data `d` to fit the transform and the return a treated training set: completely numeric, with no missing values.

```{r}
transform_design = vtreat::MultinomialOutcomeTreatment(
    var_list = setdiff(colnames(d), c('y', 'yc')),  # columns to transform
    outcome_name = 'y',                             # outcome variable
    cols_to_copy = c('y', 'yc')
)


# learn transform from data
d_prepared <-  transform_design$fit_transform(d)

# get statistics on the variables
score_frame <- transform_design$score_frame()
score_frame$recommended <- score_frame$varMoves & (score_frame$sig < 1/nrow(score_frame))
```

Note that for the training data `d`: `transform_design$fit_transform(d)` is **not** the same as `transform_design$fit(d)$transform(d)`; the second call can lead to nested model bias in some situations, and is **not** recommended.
For other, later data, not seen during transform design `transform_design$transform(o)` is an appropriate step.

Now examine the score frame, which gives information about each new variable, including its type, which original variable it is  derived from, its (cross-validated) correlation with the outcome, and its (cross-validated) significance as a one-variable linear model for the outcome. 

```{r}
knitr::kable(score_frame)
```


Note that the variable `xc` has been converted to multiple variables: 

* an indicator variable for each possible level (`xc_lev_*`)
* the value of a (cross-validated) one-variable model for `yc` as a function of `xc` (`xc_catB`)
* a variable that returns how prevalent this particular value of `xc` is in the training data (`xc_catP`)
* a variable indicating when `xc` was `NA` in the original data (`xc_lev_NA` for categorical variables, `x_isBAD` for continuous variables).

Any or all of these new variables are available for downstream modeling.

The `recommended` column indicates which variables are non constant (`varMoves` == TRUE) and have a significance value smaller than `1/nrow(score_frame)`. See the section *Deriving the Default Thresholds* below for the reasoning behind the default thresholds. Recommended columns are intended as advice about which variables appear to be most likely to be useful in a downstream model. This advice attempts to be conservative, to reduce the possibility of mistakenly eliminating variables that may in fact be useful (although, obviously, it can still mistakenly eliminate variables that have a real but non-linear relationship to the output, as is the case with `x`, in our example). Since each variable has multiple recommendations, one can consider a variable to be recommended if it is recommended for any of the outcome targets: an OR of all the recommendations.

## Examining variables

To select variables we either make our selection in terms of new variables as follows.

```{r}
good_new_variables = unique(score_frame[score_frame[['recommended']], 'varName', drop = TRUE])
good_new_variables
```

Or in terms of original variables as follows.

```{r}
good_original_variables = unique(score_frame[score_frame[['recommended']], 'origName', drop = TRUE])
good_original_variables
```

Notice, in each case we must call `unique()` as each variable (derived or original) is potentially qualified against each possible outcome.

Notice that `d_prepared` only includes derived variables and the outcome `yc`:

```{r}
d_prepared %.>%
  head(.) %.>%
  knitr::kable(.)
```

## Using the Prepared Data in a Model

Of course, what we really want to do with the prepared training data is to fit a model jointly with all the (recommended) variables. 
Let's try fitting a logistic regression model to `d_prepared`.

```{r}
library(glmnet)

model_vars <- score_frame$varName[score_frame$recommended]

model <- glmnet(x = as.matrix(d_prepared[, model_vars, drop=FALSE]), 
                y = d_prepared[['yc']],
                family = 'multinomial')
```

```{r}
# convenience functions for predicting and adding predictions to original data frame

add_predictions <- function(d_prepared, model_vars, model) {
  preds <- predict(
    model, 
    newx = as.matrix(d_prepared[, model_vars, drop=FALSE]),
    s = min(model$lambda),  # in practice we would cross-validated for a good s
    type = 'response')
  preds <- as.data.frame(preds[, , 1])
  preds$prob_on_predicted_class <- apply(preds, 1, max)
  preds$predict <- NA_character_
  for(col in colnames(preds)) {
    alter = is.na(preds$predict) & preds[[col]] >= preds$prob_on_predicted_class
    preds$predict[alter] <- col
  }
  d_prepared <- cbind(d_prepared, preds)
  d_prepared
}

add_value_by_column <- function(d_prepared, name_column, new_column) {
  vals = unique(d_prepared[[name_column]])
  d_prepared[[new_column]] <- NA_real_
  for(col in vals) {
    match <- d_prepared[[name_column]] == col
    d_prepared[[new_column]][match] <- d_prepared[match, col, drop=TRUE]
  }
  d_prepared
}
```

```{r}
# now predict
d_prepared <- add_predictions(d_prepared, model_vars, model)
d_prepared <- add_value_by_column(d_prepared, 'yc', 'prob_on_correct_class')

to_print <- c('yc', 'predict', 'large','liminal','small', 'prob_on_predicted_class', 'prob_on_correct_class')
d_prepared[, to_print, drop = FALSE] %.>%
  head(.) %.>%
  knitr::kable(.)
```

```{r}
table(truth = d_prepared$yc, prediction = d_prepared$predict)
```

In the above confusion matrix, the entry `[row, column]` gives the number of true items of `class[row]` that also have prediction of `class[column]`. In other words, the entry `['large', 'liminal']` gives the number of 'large' items predicted to be 'liminal'.

Now apply the model to new data.

```{r}
# create the new data
dtest <- make_data(450)

# prepare the new data with vtreat
dtest_prepared = transform_design$transform(dtest)

dtest_prepared <- add_predictions(dtest_prepared, model_vars, model)
dtest_prepared <- add_value_by_column(dtest_prepared, 'yc', 'prob_on_correct_class')

dtest_prepared[, to_print, drop = FALSE] %.>%
  head(.) %.>%
  knitr::kable(.)
```

```{r}
table(truth = dtest_prepared$yc, prediction = dtest_prepared$predict)
```

## Parameters for `mkCrossFrameMExperiment`

We've tried to set the defaults for all parameters so that `vtreat` is usable out of the box for most applications.


```{r}
suppressPackageStartupMessages(library(printr))
help("mkCrossFrameMExperiment")
```

Some parameters of note include:

**codeRestriction**: The types of synthetic variables that `vtreat` will (potentially) produce. See *Types of prepared variables* below.

**minFraction**: For categorical variables, indicator variables (type `indicator_code`) are only produced for levels that are present at least `minFraction` of the time. A consequence of this is that 1/`minFraction` is the maximum number of indicators that will be produced for a given categorical variable. To make sure that *all* possible indicator variables are produced, set `minFraction = 0`

**splitFunction**: The cross validation method used by `vtreat`. Most people won't have to change this.

**ncross**: The number of folds to use for cross-validation

**customCoders**: For passing in user-defined transforms for custom data preparation. Won't be needed in most situations, but see [here](http://www.win-vector.com/blog/2017/09/custom-level-coding-in-vtreat/) for an example of applying a GAM transform to input variables.


## Types of prepared variables

**clean**: Produced from numerical variables: a clean numerical variable with no `NAs` or missing values

**lev**: Produced from categorical variables, one for each (common) level: for each level of the variable, indicates if that level was "on"

**catP**: Produced from categorical variables: indicates how often each level of the variable was "on"

**catB**: Produced from categorical variables: score from a one-dimensional model of the centered output as a function of the variable

**is_BAD**: Produced for both numerical and categorical variables: an indicator variable that marks when the original variable was missing or `NaN`.

More on the coding types can be found [here](https://winvector.github.io/vtreat/articles/vtreatVariableTypes.html).

### Example: Produce only a subset of variable types

In this example, suppose you only want to use indicators and continuous variables in your model; 
in other words, you only want to use variables of types (`clean_copy`, `missing_indicator`, and `indicator_code`), and no `catB` or `prevalence_code` variables.

```{r}
transform_design_thin = vtreat::mkCrossFrameMExperiment(
    d = d,                                         # data to learn transform from
    vars = setdiff(colnames(d), c('y', 'yc')),     # columns to transform
    y_name = 'yc',                                 # outcome variable
    codeRestriction = c('lev',                     # transforms we want
                        'clean',
                        'isBAD')
)
transform_thin <- transform_design_thin$treat_m
d_prepared_thin <- transform_design_thin$cross_frame
score_frame_thin <- transform_design_thin$score_frame
score_frame_thin$recommended <- score_frame_thin$varMoves & (score_frame_thin$sig < 1/nrow(score_frame))


d_prepared_thin %.>%
  head(.) %.>%
  knitr::kable(.)
```

```{r}
knitr::kable(score_frame_thin)
```


## Deriving the Default Thresholds

While machine learning algorithms are generally tolerant to a reasonable number of irrelevant or noise variables, too many irrelevant variables can lead to serious overfit; see [this article](http://www.win-vector.com/blog/2014/02/bad-bayes-an-example-of-why-you-need-hold-out-testing/) for an extreme example, one we call "Bad Bayes". The default threshold is an attempt to eliminate obviously irrelevant variables early.

Imagine that you have a pure noise dataset, where none of the *n* inputs are related to the output. If you treat each variable as a one-variable model for the output, and look at the significances of each model, these significance-values will be uniformly distributed in the range [0:1]. You want to pick a weakest possible significance threshold that eliminates as many noise variables as possible. A moment's thought should convince you that a threshold of *1/n* allows only one variable through, in expectation. 
This leads to the general-case heuristic that a significance threshold of *1/n* on your variables should allow only one irrelevant variable through, in expectation (along with all the relevant variables). Hence, *1/n* used to be our recommended threshold, when we developed the R version of `vtreat`.

As noted above, because `vtreat` estimates variable significances using linear methods by default, some variables with a non-linear relationship  to the output may fail to pass the threshold. So it may make sense for the data scientist to filter (or not) as they will.

## Conclusion


In all cases (classification, regression, unsupervised, and multinomial classification) the intent is that `vtreat` transforms are essentially one liners.

The preparation commands are organized as follows:

 * **Regression**: [`R` regression example](https://github.com/WinVector/vtreat/blob/master/Examples/Regression/Regression.md), [`Python` regression example](https://github.com/WinVector/pyvtreat/blob/master/Examples/Regression/Regression.md).
 * **Classification**: [`R` classification example](https://github.com/WinVector/vtreat/blob/master/Examples/Classification/Classification.md), [`Python` classification  example](https://github.com/WinVector/pyvtreat/blob/master/Examples/Classification/Classification.md).
 * **Unsupervised tasks**: [`R` unsupervised example](https://github.com/WinVector/vtreat/blob/master/Examples/Unsupervised/Unsupervised.md), [`Python` unsupervised example](https://github.com/WinVector/pyvtreat/blob/master/Examples/Unsupervised/Unsupervised.md).
 * **Multinomial classification**: [`R` multinomial classification example](https://github.com/WinVector/vtreat/blob/master/Examples/Multinomial/MultinomialExample.md), [`Python` multinomial classification example](https://github.com/WinVector/pyvtreat/blob/master/Examples/Multinomial/MultinomialExample.md).

These current revisions of the examples are designed to be small, yet complete.  So as a set they have some overlap, but the user can rely mostly on a single example for a single task type.

