---
title: "Modeling Pipelines"
output: github_document
---

Reusable modeling pipelines are a practical idea that gets re-developed many times in many contexts.  [`wrapr`](https://github.com/WinVector/wrapr) supplies a particularly powerful pipeline notation, and as of version `1.8.0` pipeline re-use system (notes [here](https://winvector.github.io/wrapr/articles/Function_Objects.html)).  We will demonstrate this with the [`vtreat`](https://github.com/WinVector/vtreat) data preparation system.

```{r setup}
library("wrapr")
library("vtreat")
library("glmnet")
library("WVPlots")

ncore <- parallel::detectCores()
cl <- parallel::makeCluster(ncore)
library("doParallel")
registerDoParallel(cl)

# function to make practice data
mk_data <- function(nrows, n_var_cols, n_noise_cols) {
  d <- data.frame(y = rnorm(nrows))
  for(i in seq_len(n_var_cols)) {
    vari = paste0("var_", sprintf("%03g", i))
    d[[vari]] <- rnorm(nrows)
    d$y <- d$y + (2/n_var_cols)*d[[vari]]
    d[[vari]][d[[vari]]>abs(2*rnorm(nrows))] <- NA
    d[[vari]] <- rlnorm(1, meanlog=10, sdlog = 10)*d[[vari]]
  }
  for(i in seq_len(n_noise_cols)) {
    vari = paste0("noise_", sprintf("%03g", i))
    d[[vari]] <- rnorm(nrows)
    d[[vari]][d[[vari]]>abs(2*rnorm(nrows))] <- NA
    d[[vari]] <- rlnorm(1, meanlog=10, sdlog = 10)*d[[vari]]
  }
  d
}

set.seed(2018)
d <- mk_data(10000, 10, 200)
is_train <- runif(nrow(d))<=0.5
dTrain <- d[is_train, , drop = FALSE]
dTest <- d[!is_train, , drop = FALSE]
outcome_name <- "y"
vars <- setdiff(colnames(dTrain), outcome_name)
```

Suppose our analysis plan is the following:

* Fix missing values with `vtreat`.
* Scale and center the data.
* Model `y` as a function of the other columns using `glmnet`.

Now both `vtreat` and `glmnet` can scale, but we are going to keep the scaling
as a separate step to show how composite data preparation pipelines work.

First we combine the pre-processing steps, and a fit model as follows.

```{r model1}
# design a treatment plan using cross-validation methods
cp <- vtreat::mkCrossFrameNExperiment(
  dTrain, vars, outcome_name,
  parallelCluster = cl)

# get the list of new variables
sf <- cp$treatments$scoreFrame
newvars <- sf$varName[sf$sig <= 1/nrow(sf)]
print(newvars)

# learn a centering and scaling of the cross-validated 
# training frame
tfs <- scale(cp$crossFrame[, newvars, drop = FALSE], 
             center = TRUE, scale = TRUE)
centering <- attr(tfs, "scaled:center")
scaling <- attr(tfs, "scaled:scale")

# apply the centering and scaling to the cross-validated 
# training frame
tfs <- scale(cp$crossFrame[, newvars, drop = FALSE],
             center = centering,
             scale = scaling)

# build a cross-validation strategy to help us
# search for a good alph hyper-parameter value
cplan <- vtreat::kWayStratifiedY(
  nrow(dTrain), 5, dTrain, dTrain[[outcome_name]])
# convert the plan to cv.glmnet group notation
foldid <- numeric(nrow(dTrain))
for(i in seq_len(length(cplan))) {
  cpi <- cplan[[i]]
  foldid[cpi$app] <- i
}

# search for best cross-validated alpha
alphas <- seq(0, 1, by=0.05)
cross_scores <- vapply(
  alphas,
  function(alpha) {
    model <- cv.glmnet(as.matrix(tfs), 
                       cp$crossFrame[[outcome_name]],
                       alpha = alpha,
                       family = "gaussian", 
                       standardize = FALSE,
                       foldid = foldid, 
                       parallel = TRUE)
    index <- which(model$lambda == model$lambda.1se)[[1]]
    score <- model$cvm[[index]]
  }, numeric(1))
best_i <- which(cross_scores==min(cross_scores))[[1]]
alpha <- alphas[[best_i]]
print(alpha)

# re-fit model with chosen alpha
model <- cv.glmnet(as.matrix(tfs), 
                   cp$crossFrame[[outcome_name]],
                   alpha = alpha,
                   family = "gaussian", 
                   standardize = FALSE,
                   nfolds = 5, 
                   parallel = TRUE)
```

The question then is: how do we share such a model?  Roughly we need to share the model, any fit parameters (such as centering and scaling choices), *and* the code sequence to apply all of these steps in the proper order.

A really neat way to do this is the following.

Use `wrapr`'s ["function object" abstraction](https://winvector.github.io/wrapr/articles/Function_Objects.html), which treats names of functions, plus arguments as an efficient notation for partial evaluation.  We can use this system to encode our model prediction pipeline as follows.

```{r}
pipeline <-
  pkgfn("vtreat::prepare",
        arg_name = "dframe", 
        args = list(treatmentplan = cp$treatments,
                    varRestriction = newvars)) %.>%
  pkgfn("subset",
        arg_name = "x",
        args = list(select = newvars))  %.>%
  pkgfn("scale",
        arg_name = "x",
        args = list(center = centering,
                    scale = scaling))  %.>%
  pkgfn("glmnet::predict.cv.glmnet",
        arg_name = "newx",
        args = list(object = model,
                    s = "lambda.1se"))  %.>%
  srcfn(".[, cname, drop = TRUE]",
        arg_name = ".",
        args = list(cname = "1"))

cat(format(pipeline))
```


The pipeline is a simple list of steps (with some class annotations added).  

```{r}
pipeline@items

str(pipeline@items[[3]])
```

If you do not like pipe notation you can also build the pipeline using [`fnlist()`](https://winvector.github.io/wrapr/reference/fnlist.html) list notation.

And you can pipe data into the pipeline.

```{r}
dTrain %.>% pipeline %.>% head(.)
```
 
Or you can use a functional notation [`ApplyTo()`](https://winvector.github.io/wrapr/reference/ApplyTo.html).

```{r}
ApplyTo(pipeline, dTrain) %.>% head(.)
```


The pipeline can be saved, and contains the required parameters in lists.

```{r eval=FALSE}
saveRDS(dTrain, "dTrain.RDS")
saveRDS(pipeline, "pipeline.RDS")
```

Now the processing pipeline can be read back and used as follows.

```{r eval=FALSE}
# Fresh R session , not part of this markdown
library("wrapr")

pipeline <- readRDS("pipeline.RDS")
dTrain <- readRDS("dTrain.RDS")
dTrain %.>% pipeline %.>% head(.)
```
```{r echo=FALSE}
# simulate reading back for presentation, to make sure values match
dTrain %.>% pipeline %.>% head(.)
```

We can use this pipeline on different data, as we do to create performance plots below.

```{r}
dTrain$prediction <- dTrain %.>% pipeline

WVPlots::ScatterHist(
  dTrain, "prediction", "y", "fit on training data",
  smoothmethod = "identity",
  estimate_sig = TRUE,
  point_alpha = 0.1,
  contour = TRUE)

dTest$prediction <- dTest %.>% pipeline

WVPlots::ScatterHist(
  dTest, "prediction", "y", "fit on test",
  smoothmethod = "identity",
  estimate_sig = TRUE,
  point_alpha = 0.1,
  contour = TRUE)
```


And that is how to effectively save, share, and deploy non-trivial modeling workflows.


```{r cleanup}
parallel::stopCluster(cl)
```

