---
title: "Imputation (using fit_prepare notation)"
output: github_document
---

## Changing the missing value imputation in vtreat

For this example, we will use the `UnsupervisedTreatment`, but the same parameters can be used with the other treatment plans as well.

### A simple data example

Here we create a simple data set where the inputs have missing values.

```{r}
library(vtreat)
```


```{r}
d = data.frame(
    "x" = c(0, 1, 1000, NA),
    "w" = c(3, 6, NA, 100),
    "y" = c(0, 0, 1, 1)
)

knitr::kable(d)
```

Some of the summary statistics of `d`. We're primarily interested in the inputs `x` and `w`.

```{r}
summary(d)
```

### The default missing value imputation

By default, `vtreat` fills in missing values with the mean value of the column, and adds an advisory `*_isBAD` column to mark the location of the original missing values.

```{r}
transform_design = vtreat::UnsupervisedTreatment(
    var_list = c('x', 'w'),   # columns to transform
    cols_to_copy = 'y')  # copy the y column over

# learn treatments from data
# normally we would unpack the treatments too,
# but we don't need them for this example
unpack[d_treated = cross_frame] <- 
  fit_prepare(transform_design, d)

knitr::kable(d_treated)
```


### Changing the imputation strategy

If you do not want to use the mean to fill in missing values, you can change the imputation function globally by using the parameter `missingness_imputation` in the `UnsupervisedTreatment` parameter list. Here, we fill in missing values with the median.

```{r}
median2 <- function(x, wts) {
  median(x)
}

newparams = unsupervised_parameters(
  list(missingness_imputation = median2)
)

transform_design = vtreat::UnsupervisedTreatment(
    var_list = c('x', 'w'),   # columns to transform
     cols_to_copy = 'y',  # copy the y column over
     params = newparams
)

unpack[d_treated = cross_frame] <- 
  fit_prepare(transform_design, d)

knitr::kable(d_treated)
```


You can also use a constant value instead of a function. Here we replace missing values with the value -1.

```{r}
newparams = unsupervised_parameters(
  list(missingness_imputation = -1)
)

transform_design = vtreat::UnsupervisedTreatment(
  var_list = c('x', 'w'),   # columns to transform
  cols_to_copy = 'y',  # copy the y column over
  params = newparams
)

unpack[d_treated = cross_frame] <- 
  fit_prepare(transform_design, d)

knitr::kable(d_treated)
```

### Changing the imputation strategy per column

You can control the imputation strategy per column via the map `imputation_map`. Any column not named in the imputation map will use the imputation strategy specified by the `missingness_imputation` parameter (which is the mean by default).

Here we use the maximum value to fill in the missing values for `x` and the value `0` to fill in the missing values for `w`.


```{r}
max2 <- function(x, wts) {
  max(x)
}

transform_design = vtreat::UnsupervisedTreatment(
  var_list = c('x', 'w'),   # columns to transform
  cols_to_copy = 'y',  # copy the y column over
  imputation_map = list(
    x = max2,
    w = 0
  )
)

unpack[d_treated = cross_frame] <- 
  fit_prepare(transform_design, d)

knitr::kable(d_treated)
```

If we don't specify a column, `vtreat` looks at` missingness_imputation` (in this case, `-1`).


```{r}
newparams = unsupervised_parameters(
  list(missingness_imputation = -1)
)

transform_design = vtreat::UnsupervisedTreatment(
  var_list = c('x', 'w'),   # columns to transform
  cols_to_copy = 'y',       # copy the y column over
  params = newparams,       # default missingness value
  imputation_map = list(    # custom imputations
    x = max2
  )
)

unpack[d_treated = cross_frame] <- 
  fit_prepare(transform_design, d)

knitr::kable(d_treated)
```

If `missingness_imputation` is not specified, `vtreat` uses a weighted mean.

```{r}
transform_design = vtreat::UnsupervisedTreatment(
  var_list = c('x', 'w'),   # columns to transform
  cols_to_copy = 'y',       # copy the y column over
  imputation_map = list(    # custom imputations
    x = max2
  )
)

unpack[d_treated = cross_frame] <- 
  fit_prepare(transform_design, d)
knitr::kable(d_treated)
```