---
title: "Imputation"
output: github_document
---

## Changing the missing value imputation in vtreat

For this example, we will use the UnsupervisedTreatment, but the same parameters can be used with the other treatment plans as well.

### A simple data example

Here we create a simple data set where the inputs have missing values.

```{r}
library(vtreat)
```


```{r}
d = data.frame(
    "x" = c(0, 1, 1000, NA),
    "w" = c(3, 6, NA, 100),
    "y" = c(0, 0, 1, 1)
)

knitr::kable(d)
```

Some of the summary statistics of `d`. We're primarily interested in the inputs `x` and `w`.

```{r}
summary(d)
```

### The default missing value imputation

By default, vtreat fills in missing values with the mean value of the column, and adds an advisory *_is_bad column to mark the location of the original missing values.

```{r}
treatments <- designTreatmentsZ(d, 
                                varlist = c('x', 'w'), 
                                verbose = FALSE)
d_treated <- prepare(treatments, 
                     d)
d_treated$y <- d$y
knitr::kable(d_treated)
```


### Changing the imputation strategy

If you do not want to use the mean to fill in missing values, you can change the imputation function using the parameter missingness_imputation. Here, we fill in missing values with the median.

```{r}
median2 <- function(x, wts) {
  median(x)
}

treatments <- designTreatmentsZ(d, 
                                varlist = c('x', 'w'), 
                                verbose = FALSE,
                                missingness_imputation = median2)
d_treated <- prepare(treatments, 
                     d)
d_treated$y <- d$y
knitr::kable(d_treated)
```


You can also use a constant value instead of a function. Here we replace missing values with the value -1.

```{r}
treatments <- designTreatmentsZ(d, 
                                varlist = c('x', 'w'), 
                                verbose = FALSE,
                                missingness_imputation = -1)
d_treated <- prepare(treatments, 
                     d)
d_treated$y <- d$y
knitr::kable(d_treated)
```


### Changing the imputation strategy per column

You can control the imputation strategy per column via the map imputation_map. Any column not named in the imputation map will use the imputation strategy specified by the missingness_imputation parameter (which is the mean by default).

Here we use the maximum value to fill in the missing values for `x` and the value `0` to fill in the missing values for `w`.

```{r}
max2 <- function(x, wts) {
  max(x)
}

treatments <- designTreatmentsZ(d, 
                                varlist = c('x', 'w'), 
                                verbose = FALSE,
                                imputation_map = list(
                                  x = max2,
                                  w = 0
                                ))
d_treated <- prepare(treatments, 
                     d)
d_treated$y <- d$y
knitr::kable(d_treated)
```




If we don't specify a column, `vtreat` looks at` missingness_imputation` (in this case, `-1`).

```{r}
treatments <- designTreatmentsZ(d, 
                                varlist = c('x', 'w'), 
                                verbose = FALSE,
                                missingness_imputation = -1,
                                imputation_map = list(
                                  x = max2
                                ))
d_treated <- prepare(treatments, 
                     d)
d_treated$y <- d$y
knitr::kable(d_treated)
```


If `missingness_imputation` is not specified, vtreat uses a weighted mean.

```{r}
treatments <- designTreatmentsZ(d, 
                                varlist = c('x', 'w'), 
                                verbose = FALSE,
                                imputation_map = list(
                                  x = max2
                                ))
d_treated <- prepare(treatments, 
                     d)
d_treated$y <- d$y
knitr::kable(d_treated)
```
